---
title: "Social Network Analysis of the Canadian Disability Rights Community"
subtitle: "Mapping Collaboration Networks and Key Actors in Disability Rights Advocacy"
author: "Dr. Derrick L. Cogburn"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    number-sections: true
    fig-width: 12
    fig-height: 8
    embed-resources: true
  pdf:
    pdf-engine: lualatex
    keep-tex: true
    titlepage: true
    toc: true
    toc-depth: 2
    code-block-wrap: true
    number-sections: true
    include-in-header:
      text: |
        \usepackage{xurl} % break long URLs
        \PassOptionsToPackage{hyphens}{url}
        \usepackage{fvextra} % better verbatim handling (no Highlighting redefinition)
execute:
  echo: true
  warning: false
  message: false
---

# Setup

## Install Required Packages

Run this chunk once to install all dependencies. After installation, you can set
`eval: false` or simply skip it on future renders.

```{r install-packages}
#| eval: false

install.packages(c(
  "tidyverse",      # Data wrangling and ggplot2

"igraph",          # Core network analysis
  "tidygraph",      # Tidy interface to igraph
  "ggraph",         # Network visualisation (ggplot2-based)
  "readxl",         # Read .xlsx files
  "patchwork",      # Combine ggplot panels
  "scales",         # Axis formatting
  "RColorBrewer",   # Colour palettes
  "knitr",          # Tables in Quarto
  "kableExtra"      # Styled tables
))
```

## Load Libraries

```{r load-libraries}
library(tidyverse)
library(igraph)
library(tidygraph)
library(ggraph)
library(readxl)
library(patchwork)
library(scales)
library(RColorBrewer)
library(knitr)
library(kableExtra)

theme_set(theme_minimal(base_size = 12))
set.seed(42)
```


```{r setup}
#| include: false
library(tidyverse)
library(igraph)
library(tidygraph)
library(ggraph)
library(readxl)
library(janitor)

# NOTE: Update this path to your local file location
data_file <- "Enque_te_d_analyse_des_Re_seaux_Sociaux_-_Social_Network_Analysis_Survey_-_combined_Mar252026-_deidentified.xlsx"
```


```{r data-prep}
#| include: false

# Read and clean data - you may need to adjust column references
# The following uses the pre-cleaned CSVs from the Python pipeline.
# Alternatively, you can read the raw xlsx and replicate the cleaning in R.

# If using the pre-cleaned files:
respondents <- read_csv("respondents_final.csv")
edges <- read_csv("edges_final.csv")
```

## Two-Mode Network: Respondents and Issue Areas

```{r bipartite-network}
#| fig-cap: "Two-mode bipartite network. Circle nodes represent respondents; square nodes represent issue areas."

# Build bipartite graph
bip_edges <- respondents |>
  filter(!is.na(issue_category), issue_category != "Unknown") |>
  separate_rows(issue_category, sep = ";") |>
  select(resp_id, issue_category) |>
  filter(issue_category != "Unknown", nchar(issue_category) > 0)

bip_graph <- graph_from_data_frame(bip_edges, directed = FALSE)
V(bip_graph)$type <- V(bip_graph)$name %in% unique(bip_edges$issue_category)

as_tbl_graph(bip_graph) |>
  ggraph(layout = "fr") +
  geom_edge_link(alpha = 0.08, colour = "#666666") +
  geom_node_point(aes(shape = type, size = centrality_degree(),
                      colour = type), alpha = 0.7) +
  geom_node_text(aes(label = ifelse(type, name, NA)),
                 size = 3, fontface = "bold", repel = TRUE) +
  scale_shape_manual(values = c(16, 15),
                     labels = c("Respondent", "Issue Area")) +
  scale_colour_manual(values = c("#457B9D", "#E63946"),
                      labels = c("Respondent", "Issue Area")) +
  scale_size_continuous(range = c(2, 12)) +
  theme_void() +
  labs(title = "Bipartite Network: Respondents and Disability Rights Issue Areas") +
  theme(legend.position = "bottom")
```

## One-Mode Projected Network

```{r projected-network}
#| fig-cap: "Respondents linked by shared collaboration partners."

# Build respondent-org bipartite
resp_org <- edges |>
  filter(!is.na(partner_clean), partner_clean != "nan") |>
  select(respondent, partner_clean)

bip_ro <- graph_from_data_frame(resp_org, directed = FALSE)
V(bip_ro)$type <- !(V(bip_ro)$name %in% unique(resp_org$respondent))

# Project onto respondent mode
proj <- bipartite_projection(bip_ro, which = "false")

# Add attributes
resp_lookup <- respondents |> select(resp_id, primary_connection, primary_issue)
V(proj)$connection <- resp_lookup$primary_connection[match(V(proj)$name, resp_lookup$resp_id)]

as_tbl_graph(proj) |>
  ggraph(layout = "fr") +
  geom_edge_link(aes(width = weight), alpha = 0.3, colour = "#999999") +
  geom_node_point(aes(colour = connection, size = centrality_degree()), alpha = 0.8) +
  scale_edge_width(range = c(0.5, 3)) +
  theme_void() +
  labs(title = "Projected Network: Respondents Linked by Shared Partners",
       colour = "Connection Type") +
  theme(legend.position = "bottom")
```

## Centrality Analysis

```{r centrality}
#| fig-cap: "Organization degree centrality — most frequently named collaboration partners."

# Organization degree
org_deg <- resp_org |>
  count(partner_clean, sort = TRUE) |>
  head(15)

ggplot(org_deg, aes(x = reorder(partner_clean, n), y = n)) +
  geom_col(fill = "#457B9D", alpha = 0.85) +
  coord_flip() +
  labs(x = NULL, y = "Degree (respondents naming this org)",
       title = "Organization Degree Centrality") +
  theme_minimal()
```

## Community Detection

```{r communities}
#| fig-cap: "Louvain community detection on the projected network."

if (ecount(proj) > 0) {
  comms <- cluster_louvain(proj, weights = E(proj)$weight)

  as_tbl_graph(proj) |>
    mutate(community = as.factor(membership(comms))) |>
    ggraph(layout = "fr") +
    geom_edge_link(alpha = 0.15, colour = "#BBBBBB") +
    geom_node_point(aes(colour = community, size = centrality_degree()), alpha = 0.85) +
    theme_void() +
    labs(title = paste0("Community Detection (Louvain): ",
                        length(unique(membership(comms))), " Communities"),
         colour = "Community") +
    theme(legend.position = "bottom")
}
```

## Degree Distributions

```{r distributions}
#| fig-cap: "Degree distributions and collaboration frequency."
#| fig-width: 16
#| fig-height: 5

p1 <- resp_org |>
  count(partner_clean) |>
  ggplot(aes(x = n)) +
  geom_histogram(binwidth = 1, fill = "#457B9D", alpha = 0.8, colour = "white") +
  labs(x = "Degree", y = "Count", title = "Organization Degree Distribution") +
  theme_minimal()

freq_data <- edges |>
  filter(freq_clean != "Unknown") |>
  count(freq_clean)

p2 <- ggplot(freq_data, aes(x = freq_clean, y = n, fill = freq_clean)) +
  geom_col(alpha = 0.85) +
  scale_fill_manual(values = c("Regularly" = "#E63946",
                                "Occasionally" = "#E9C46A",
                                "Seldom" = "#457B9D")) +
  labs(x = NULL, y = "Count", title = "Collaboration Frequency") +
  theme_minimal() +
  theme(legend.position = "none")

library(patchwork)
p1 + p2
```

## Discussion and Implications

The Canadian disability rights community operates as a **highly decentralized network**. Collaboration is predominantly local and issue-specific rather than nationally integrated. A small number of national organizations serve as critical connectors whose capacity to convene and bridge across issue silos is disproportionately important to overall network cohesion.
