---
title: "Social Network Analysis of the Canadian Disability Rights Community"
subtitle: "Mapping Collaboration Networks and Key Actors in Disability Rights Advocacy"
author: "Dr. Derrick L. Cogburn"
date: today
format:
  pdf:
    pdf-engine: lualatex
    keep-tex: true
    titlepage: true
    toc: true
    toc-depth: 2
    code-block-wrap: true
    number-sections: true
    fig-width: 6.5
    fig-height: 4.5
    include-in-header:
      text: |
        \usepackage{xurl} % break long URLs
        \PassOptionsToPackage{hyphens}{url}
        \usepackage{fvextra} % better verbatim handling (no Highlighting redefinition)
        \usepackage{booktabs} % professional table rules
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    number-sections: true
    fig-width: 12
    fig-height: 8
    embed-resources: true
    code-fold: true
    code-summary: "Show code"
execute:
  echo: true
  warning: false
  message: false
---

# Executive Summary

This report presents a Social Network Analysis (SNA) of collaboration patterns within the Canadian disability rights community, based on a bilingual survey (English/French). The survey, called "Enquête d'analyse des Réseaux Sociaux / Social Network Analysis Survey" enables an analysis and mapping of who collaborates with whom, identifies the most central actors and organizations, detects community clusters, and assesses overall network structure.

# Data Cleaning and Pipeline

## Source Data

The raw data was collected via a bilingual (English/French) Microsoft Forms survey distributed to members of the Canadian disability rights community. The survey instrument contained 51 columns across a single worksheet, with parallel English and French versions of each question. A total of 111 responses were recorded. The raw data file was de-identified prior to analysis.

## Survey Structure and Key Fields

The survey captured several categories of information relevant to social network analysis. Respondent attributes included the respondent's connection to the disability community (e.g., person with a disability, family member/caregiver, advocate, researcher, health professional) and the disability rights issues in which they are most actively involved (free-text responses covering topics such as employment, children/youth, housing, poverty, and accessibility). Collaboration data included the names of up to three organizations or individuals with whom the respondent has partnered in the past five years (free-text), the frequency of collaboration with each named partner (regularly, occasionally, or seldom), and whether the respondent's named partners are linked to each other.

## Data Cleaning Steps

The cleaning pipeline proceeded through several stages. First, the metadata row generated by Microsoft Forms was removed, and the language preference field was used to route each respondent to the appropriate English or French column for each question. Second, the bilingual fields were consolidated, with the French-language response used when the respondent selected French and the English-language response used otherwise. This produced a unified respondent table with consistent column names regardless of survey language.

Third, connection types were standardized into nine categories: Person with Disability, Advocate/CSO, Family/Caregiver, Researcher, Health, Education, Leisure, Government/Policy, and Other. Respondents whose connection type text matched multiple categories were assigned all applicable categories (semicolon-separated), with the first match designated as the primary connection type.

Fourth, disability rights issue areas were categorized from the free-text responses into 12 standardized topics: Children/Youth, Employment, Accessibility/Inclusion, Poverty/Income, Rights/Policy, Education, Housing, Health, Indigenous, Transportation, Technology, and Other. As with connection types, respondents active in multiple areas received all applicable categories.

Fifth, the free-text partner names were parsed into individual organization or person names. The parser handled several formats including comma-separated lists, semicolon-separated lists, numbered lists, and narrative descriptions. Each respondent contributed up to three partner names. Names were then normalized by converting to lowercase, collapsing whitespace, and removing punctuation. A prefix-matching algorithm was applied to merge near-duplicate entries (e.g., variant spellings or abbreviations of the same organization). This best-effort automated cleaning reduced the number of unique partner names from 175 to 166.

Sixth, collaboration frequency values were harmonized across languages, mapping both English terms (Regularly, Occasionally, Seldom) and their French equivalents (Régulièrement, Occasionnellement, Rarement) to a common three-level scale plus an Unknown category for non-standard responses.

## Output Files

The pipeline produced two analysis-ready CSV files. The node table (`respondents_final.csv`) contains one row per respondent with fields for respondent ID, language, raw and categorized connection type, raw and categorized issue areas, and primary connection and issue assignments. The edge table (`edges_final.csv`) contains one row per respondent-partner tie with fields for respondent ID, raw partner name, normalized partner name, cleaned canonical partner name, raw frequency, and harmonized frequency.

# Setup

## Install Required Packages

Run this chunk once to install all dependencies. After installation, you can set
`eval: false` or simply skip it on future renders.

```{r}
#| label: install-packages
#| eval: false

install.packages(c(
  "tidyverse",      # Data wrangling and ggplot2
  "igraph",         # Core network analysis
  "tidygraph",      # Tidy interface to igraph
  "ggraph",         # Network visualisation (ggplot2-based)
  "readxl",         # Read .xlsx files
  "patchwork",      # Combine ggplot panels
  "scales",         # Axis formatting
  "RColorBrewer",   # Colour palettes
  "knitr",          # Tables in Quarto
  "kableExtra"      # Styled tables
))
```

## Load Libraries

```{r}
#| label: load-libraries
library(tidyverse)
library(igraph)
library(tidygraph)
library(ggraph)
library(readxl)
library(patchwork)
library(scales)
library(RColorBrewer)
library(knitr)
library(kableExtra)

theme_set(theme_minimal(base_size = 12))
set.seed(42)
```

## Import Data

The analysis uses two pre-cleaned CSV files produced by Dr. Cogburn in a data-preparation pipeline. These files, `respondents_final.csv` and `edges_final.csv` are located in the data folder, which is located in the same directory as this `.qmd` file. Paths to the data should be updated to data/filename.

```{r}
#| label: import-data
# ── Node table: one row per survey respondent ──
respondents <- read_csv("data/respondents_final.csv", show_col_types = FALSE) |>
  # Drop the metadata row carried over from MS Forms
filter(resp_id != "RMicrosoft Forms ")

# ── Edge table: one row per respondent → partner tie ──
edges <- read_csv("data/edges_final.csv", show_col_types = FALSE) |>
  filter(
    !is.na(partner_clean),
    partner_clean != "nan",
    partner_clean != "None"
  )
```

```{r}
#| label: quick-summary
cat(
  "Respondents:", nrow(respondents), "\n",
  "Respondents who named partners:", n_distinct(edges$respondent), "\n",
  "Unique organisations named:", n_distinct(edges$partner_clean), "\n",
  "Total collaboration ties:", nrow(edges), "\n"
)
```

## Colour Palettes

```{r}
#| label: palettes
# Connection-type palette (respondent node colouring)
conn_palette <- c(
  "Person with Disability" = "#E63946",
  "Advocate/CSO"           = "#457B9D",
  "Family/Caregiver"       = "#2A9D8F",
  "Researcher"             = "#6A4C93",
  "Health"                 = "#BC4749",
  "Education"              = "#F4A261",
  "Leisure"                = "#8AC926",
  "Government/Policy"      = "#264653",
  "Other"                  = "#AAAAAA",
  "Unknown"                = "#BBBBBB"
)

# Issue-area palette (bipartite squares)
issue_palette <- c(
  "Children/Youth"          = "#E63946",
  "Employment"              = "#457B9D",
  "Accessibility/Inclusion" = "#2A9D8F",
  "Poverty/Income"          = "#E9C46A",
  "Rights/Policy"           = "#264653",
  "Education"               = "#F4A261",
  "Housing"                 = "#606C38",
  "Health"                  = "#BC4749",
  "Indigenous"              = "#6A4C93",
  "Transportation"          = "#1982C4",
  "Technology"              = "#8AC926",
  "Other"                   = "#AAAAAA"
)

# Collaboration-frequency palette
freq_palette <- c(
  "Regularly"    = "#E63946",
  "Occasionally" = "#E9C46A",
  "Seldom"       = "#457B9D",
  "Unknown"      = "#BBBBBB"
)
```


# Two-Mode Bipartite Network: Respondents × Issue Areas

This network maps respondents (circles) to the disability rights issue areas they are most actively involved in (squares). Respondents are coloured by their connection type to the disability community; issue-area nodes are sized by the number of affiliated respondents.

```{r}
#| label: bipartite-build

# Expand the semicolon-separated issue_category into long form
bip_edges <- respondents |>
  filter(
    !is.na(issue_category),
    issue_category != "Unknown"
  ) |>
  separate_rows(issue_category, sep = ";") |>
  mutate(issue_category = str_trim(issue_category)) |>
  filter(issue_category != "", issue_category != "Unknown") |>
  select(resp_id, issue_category)

# Build igraph bipartite
bip_graph <- graph_from_data_frame(bip_edges, directed = FALSE)

# Mark node type: TRUE = issue-area node, FALSE = respondent node
V(bip_graph)$type <- V(bip_graph)$name %in% unique(bip_edges$issue_category)

# Attach respondent attributes
resp_lookup <- respondents |> select(resp_id, primary_connection)
V(bip_graph)$primary_connection <- resp_lookup$primary_connection[
  match(V(bip_graph)$name, resp_lookup$resp_id)
]
V(bip_graph)$primary_connection[is.na(V(bip_graph)$primary_connection)] <- "Issue Area"

# Assign each node its display colour for a unified colour scale.
# Respondents get their connection-type colour; issue-area nodes get
# their own distinct colour from the issue_palette.
V(bip_graph)$node_colour <- ifelse(
  V(bip_graph)$type,
  V(bip_graph)$name,
  V(bip_graph)$primary_connection
)
```

```{r}
#| label: bipartite-plot
#| fig-cap: "Figure 1. Two-mode bipartite network. Circles = respondents (coloured by connection type); squares = issue areas (sized by degree)."
#| fig-height: 10
#| fig-width: 10
#| out-width: "100%"

# Combined palette: respondent connection types + issue-area names
bipartite_palette <- c(conn_palette, issue_palette)

# Build the tidy graph with layout attributes
bip_tbl <- as_tbl_graph(bip_graph) |>
  mutate(
    node_type   = ifelse(type, "Issue Area", "Respondent"),
    node_degree = centrality_degree(),
    # Respondent nodes: small, scaled by degree
    # Issue-area nodes: larger, scaled more aggressively so they stand out
    node_size = ifelse(type, node_degree * 0.6 + 4, node_degree * 0.4 + 1.5)
  )

ggraph(bip_tbl, layout = "fr", niter = 800) +
  # Edges: very light so the structure is visible without clutter
  geom_edge_link(alpha = 0.06, colour = "#666666", width = 0.3) +
  # Respondent nodes: circles, colour mapped for legend
  geom_node_point(
    data = . %>% filter(!type),
    aes(colour = node_colour, size = node_size),
    shape = 16, alpha = 0.7
  ) +
  # Issue-area nodes: filled squares with dark border, drawn on top
  geom_node_point(
    data = . %>% filter(type),
    aes(fill = node_colour, size = node_size),
    shape = 22, colour = "#333333", stroke = 0.8, alpha = 0.85
  ) +
  # Labels only on issue-area nodes, repelled for readability
  geom_node_text(
    aes(label = ifelse(type, name, NA)),
    size = 2.8, fontface = "bold", colour = "#111111",
    repel = TRUE, max.overlaps = 20,
    point.padding = unit(0.4, "lines"),
    box.padding = unit(0.5, "lines"),
    segment.colour = "#999999", segment.size = 0.3
  ) +
  # Respondent legend via colour scale
  scale_colour_manual(
    values = conn_palette,
    name   = "Respondent Connection Type",
    guide  = guide_legend(
      override.aes = list(size = 3, shape = 16, alpha = 0.8)
    )
  ) +
  # Issue-area squares: no legend (they are self-labelled)
  scale_fill_manual(values = issue_palette, guide = "none") +
  scale_size_identity() +
  labs(title = "Bipartite Network: Respondents and Disability Rights Issue Areas") +
  theme_void() +
  theme(
    plot.title      = element_text(face = "bold", size = 13, hjust = 0.5),
    legend.position = "bottom",
    legend.title    = element_text(size = 8, face = "bold"),
    legend.text     = element_text(size = 7),
    plot.margin     = margin(5, 5, 5, 5)
  )
```


# One-Mode Projected Network: Respondent Collaboration

Two respondents are linked in the projected network when they named the same
organisation as a collaboration partner. Edge weight reflects the number of
shared partners.

```{r}
#| label: projected-build
# Build the respondent–organisation bipartite graph
resp_org_edges <- edges |>
  select(respondent, partner_clean) |>
  distinct()

bip_ro <- graph_from_data_frame(resp_org_edges, directed = FALSE)
V(bip_ro)$type <- !(V(bip_ro)$name %in% unique(resp_org_edges$respondent))

# Project onto respondent mode
proj <- bipartite_projection(bip_ro, which = "false")

# Attach respondent attributes
V(proj)$primary_connection <- resp_lookup$primary_connection[
  match(V(proj)$name, resp_lookup$resp_id)
]
V(proj)$primary_connection[is.na(V(proj)$primary_connection)] <- "Unknown"

issue_lookup <- respondents |> select(resp_id, primary_issue)
V(proj)$primary_issue <- issue_lookup$primary_issue[
  match(V(proj)$name, issue_lookup$resp_id)
]
```

```{r}
#| label: projected-plot
#| fig-cap: "Figure 2. One-mode projected network. Node size = degree; colour = connection type; edge width = number of shared partners."
#| fig-height: 10

as_tbl_graph(proj) |>
  mutate(deg = centrality_degree()) |>
  ggraph(layout = "fr") +
  geom_edge_link(
    aes(width = weight),
    alpha = 0.25, colour = "#999999"
  ) +
  geom_node_point(
    aes(colour = primary_connection, size = deg),
    alpha = 0.8
  ) +
  geom_node_text(
    aes(label = ifelse(deg >= 2, name, NA)),
    size = 2.5, repel = TRUE, max.overlaps = 15
  ) +
  scale_edge_width(range = c(0.5, 3), name = "Shared Partners") +
  scale_colour_manual(values = conn_palette, name = "Connection Type") +
  scale_size_continuous(range = c(3, 14), name = "Degree") +
  labs(title = "Projected Network: Respondents Linked by Shared Collaboration Partners") +
  theme_void() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14)
  )
```

## Network Structure Metrics

```{r}
#| label: network-metrics
metrics_tbl <- tibble(
  Metric = c(
    "Nodes (respondents with ties)",
    "Edges (shared-partner links)",
    "Density",
    "Connected components",
    "Transitivity (global clustering)",
    "Average clustering coefficient",
    "Largest component size"
  ),
  Value = c(
    vcount(proj),
    ecount(proj),
    round(graph.density(proj), 4),
    components(proj)$no,
    round(transitivity(proj, type = "global"), 4),
    round(transitivity(proj, type = "average"), 4),
    max(components(proj)$csize)
  )
)

metrics_tbl |>
  kable(align = c("l", "r"), booktabs = TRUE) |>
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    latex_options = c("hold_position"),
    full_width = FALSE
  )
```


# Centrality Analysis

Centrality measures identify the most influential actors. **Degree centrality**
counts direct connections; **betweenness centrality** identifies actors who
bridge otherwise disconnected parts of the network.

## Organisation Degree Centrality

```{r}
#| label: org-degree
#| fig-cap: "Figure 3a. The 15 most frequently named collaboration partners."
#| fig-height: 6

org_degree <- edges |>
  count(partner_clean, sort = TRUE, name = "degree") |>
  head(15)

ggplot(org_degree, aes(x = reorder(partner_clean, degree), y = degree)) +
  geom_col(fill = "#457B9D", alpha = 0.85, width = 0.7) +
  geom_text(aes(label = degree), hjust = -0.3, size = 3.5) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Degree (respondents naming this organisation)",
    title = "Organisation Degree Centrality"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

## Respondent Centrality (Projected Network)

```{r}
#| label: respondent-centrality
#| fig-cap: "Figure 3b. Respondent degree and betweenness centrality in the projected network."
#| fig-height: 7
#| fig-width: 10
#| out-width: "100%"

# Compute centrality measures
cent_df <- tibble(
  name              = V(proj)$name,
  primary_connection = V(proj)$primary_connection,
  degree            = degree(proj),
  betweenness       = betweenness(proj, weights = E(proj)$weight)
) |>
  arrange(desc(degree))

# Degree panel
p_deg <- cent_df |>
  head(12) |>
  mutate(label = paste0(name, " (", str_sub(primary_connection, 1, 12), ")")) |>
  ggplot(aes(x = reorder(label, degree), y = degree, fill = primary_connection)) +
  geom_col(alpha = 0.85, width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = conn_palette) +
  labs(x = NULL, y = "Degree", title = "Respondent Degree Centrality") +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold")
  )

# Betweenness panel
p_betw <- cent_df |>
  arrange(desc(betweenness)) |>
  head(12) |>
  mutate(label = paste0(name, " (", str_sub(primary_connection, 1, 12), ")")) |>
  ggplot(aes(x = reorder(label, betweenness), y = betweenness, fill = primary_connection)) +
  geom_col(alpha = 0.85, width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = conn_palette) +
  labs(x = NULL, y = "Betweenness", title = "Respondent Betweenness Centrality") +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold")
  )

p_deg + p_betw +
  plot_annotation(
    title = "Centrality Analysis: Key Actors in the Projected Network",
    theme = theme(plot.title = element_text(face = "bold", size = 14))
  )
```


# Community Detection

The Louvain algorithm partitions the network into communities that maximise
modularity — groups of respondents who are more densely connected to each other
than to the rest of the network.

```{r}
#| label: community-detection
#| fig-cap: "Figure 4. Louvain community detection on the projected network."
#| fig-height: 10

comms <- cluster_louvain(proj, weights = E(proj)$weight)

# Summary stats
n_comms    <- length(unique(membership(comms)))
mod_score  <- round(modularity(comms), 4)
comm_sizes <- sort(table(membership(comms)), decreasing = TRUE)

cat(
  "Communities detected:", n_comms, "\n",
  "Modularity:", mod_score, "\n",
  "Sizes:", paste(comm_sizes, collapse = ", "), "\n"
)
```

```{r}
#| label: community-plot
#| fig-cap: "Figure 4. Louvain communities. Colour = community membership; size = degree."
#| fig-height: 10

as_tbl_graph(proj) |>
  mutate(
    community = as.factor(membership(comms)),
    deg       = centrality_degree()
  ) |>
  ggraph(layout = "fr") +
  geom_edge_link(alpha = 0.15, colour = "#BBBBBB", width = 0.5) +
  geom_node_point(
    aes(colour = community, size = deg),
    alpha = 0.85
  ) +
  geom_node_text(
    aes(label = ifelse(deg >= 2, name, NA)),
    size = 2.5, repel = TRUE, max.overlaps = 15
  ) +
  scale_size_continuous(range = c(3, 14), name = "Degree") +
  labs(
    title   = paste0("Community Detection (Louvain): ", n_comms, " Communities"),
    subtitle = paste0("Modularity = ", mod_score),
    colour  = "Community"
  ) +
  theme_void() +
  theme(
    legend.position = "bottom",
    plot.title    = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11, colour = "grey40")
  )
```

## Community Composition

```{r}
#| label: community-table
# Build composition summary for multi-member communities
comm_membership <- tibble(
  name              = V(proj)$name,
  primary_connection = V(proj)$primary_connection,
  primary_issue     = V(proj)$primary_issue,
  community         = membership(comms)
)

comm_summary <- comm_membership |>
  group_by(community) |>
  summarise(
    Size = n(),
    `Connection Types` = paste(
      names(sort(table(primary_connection), decreasing = TRUE)[1:min(3, n())]),
      collapse = ", "
    ),
    `Issue Areas` = paste(
      names(sort(table(primary_issue), decreasing = TRUE)[1:min(3, n())]),
      collapse = ", "
    ),
    .groups = "drop"
  ) |>
  filter(Size >= 2) |>
  arrange(desc(Size)) |>
  mutate(Community = paste0("Community ", row_number())) |>
  select(Community, Size, `Connection Types`, `Issue Areas`)

comm_summary |>
  kable(booktabs = TRUE) |>
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    latex_options = c("hold_position", "scale_down"),
    full_width = FALSE
  )
```


# Degree Distributions and Collaboration Patterns

```{r}
#| label: degree-distributions
#| fig-cap: "Figure 5. Left: Organisation degree distribution. Centre: Respondent degree distribution (projected). Right: Collaboration frequency."
#| fig-height: 5
#| fig-width: 10
#| out-width: "100%"

# Organisation degree distribution
p1 <- edges |>
  count(partner_clean, name = "degree") |>
  ggplot(aes(x = degree)) +
  geom_histogram(
    binwidth = 1, fill = "#457B9D", alpha = 0.8,
    colour = "white", boundary = 0.5
  ) +
  geom_vline(
    aes(xintercept = mean(degree)),
    linetype = "dashed", colour = "#E63946"
  ) +
  labs(
    x = "Degree", y = "Count",
    title = "Organisation Degree Distribution"
  ) +
  theme(plot.title = element_text(face = "bold", size = 10))

# Respondent degree distribution (projected network)
resp_deg_df <- tibble(degree = degree(proj))

p2 <- ggplot(resp_deg_df, aes(x = degree)) +
  geom_histogram(
    binwidth = 1, fill = "#2A9D8F", alpha = 0.8,
    colour = "white", boundary = -0.5
  ) +
  geom_vline(
    aes(xintercept = mean(degree)),
    linetype = "dashed", colour = "#E63946"
  ) +
  labs(
    x = "Degree", y = "Count",
    title = "Respondent Degree Distribution\n(Projected Network)"
  ) +
  theme(plot.title = element_text(face = "bold", size = 10))

# Collaboration frequency
p3 <- edges |>
  filter(freq_clean != "Unknown") |>
  count(freq_clean) |>
  ggplot(aes(x = freq_clean, y = n, fill = freq_clean)) +
  geom_col(alpha = 0.85, width = 0.7) +
  geom_text(aes(label = n), vjust = -0.4, fontface = "bold", size = 3.5) +
  scale_fill_manual(values = freq_palette) +
  labs(x = NULL, y = "Count", title = "Collaboration Frequency") +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 10)
  )

p1 + p2 + p3 +
  plot_annotation(
    title = "Network Structure: Degree Distributions and Collaboration Patterns",
    theme = theme(plot.title = element_text(face = "bold", size = 14))
  )
```


# Discussion and Implications

The Canadian disability rights community operates as a **highly decentralized network**. Collaboration is predominantly local and issue-specific rather than nationally integrated. A small number of national organizations serve as critical connectors whose capacity to convene and bridge across issue silos is disproportionately important to overall network cohesion.

## Fragmentation as a structural feature

The Canadian disability rights community, as captured by this survey, operates
as a highly decentralised network. Collaboration is predominantly local and
issue-specific rather than nationally integrated. This is not necessarily a
weakness — it may reflect the diversity of disability issues and the importance
of place-based advocacy — but it does mean that information, resources, and
strategic coordination may not flow easily across the network.

## The role of national organisations

A small number of national organisations (Disability Without Poverty, CNIB,
March of Dimes, Inclusion Canada) serve as critical connectors. These
organisations' capacity to convene, share information, and bridge across issue
silos is disproportionately important to overall network cohesion. Strengthening
these hub organisations — or creating new ones — could substantially improve
coordination across the community.

## Bridging actors

Respondents with high betweenness centrality are structurally important: they
connect otherwise isolated parts of the network. Supporting and amplifying these
bridging actors — many of whom are persons with disabilities themselves — could
enhance cross-pollination of ideas and strategies.

## Limitations

This analysis is based on a convenience sample of `r nrow(respondents)`
respondents. The free-text partner naming format means some organisations may be
inconsistently recorded. The projected one-mode network relies on the assumption
that shared organisational partnerships indicate a meaningful indirect connection
between respondents. Finally, the survey captures reported collaboration, which
may differ from actual interaction patterns.

---

*Report generated from the Enquête d'analyse des Réseaux Sociaux / Social
Network Analysis Survey. Analysis performed in R using igraph, tidygraph, and
ggraph. All respondent identifiers are de-identified survey IDs.*
